---
title: "Phase 2: Model Planning and Building"
author: "Brendan Allen"
output:
  html_document:
    df_print: paged
---

<style>
body {
text-align: justify}
</style>

####[Home](https://ballen2024.github.io/DataScience/index.html)

####[Phase 1: Discovery and Data Preparation](https://ballen2024.github.io/DataScience/deliverable1.html)

####[Phase 2: Model Planning and Building](https://ballen2024.github.io/DataScience/deliverable2.html)

####[Phase 3: Results and Operationalization](https://ballen2024.github.io/DataScience/deliverable3.html)

###Getting Started
**First we want to import any necessary libraries we will be using.**

**Note: Uncomment the install lines if you haven't yet installed these packages**
```{r warning=FALSE}
suppressMessages(library(tidyr))
suppressMessages(library(dplyr))
# suppressMessages(install.packages("devtools"))
# suppressMessages(devtools::install_github("stefanedwards/lemon"))
suppressMessages(library(lemon))
suppressMessages(library(scales))
suppressMessages(library(tidyverse))
suppressMessages(library(knitr))
knit_print.data.frame <- lemon_print
```

**We will want to import all of our work from [Phase 1](https://ballen2024.github.io/DataScience/deliverable1.html) as well.**
```{r  message=FALSE, error=FALSE, warning=FALSE, results='hide'}
purl("deliverable1.Rmd", output = "deliverable1.r")
source("deliverable1.r")
```

**Let's read in our new data on foul balls**
```{r}
foul_balls <- suppressMessages(read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/foul-balls/foul-balls.csv"))
```

###Looking at the Data
The data source [*Baseball Savant**](https://baseballsavant.mlb.com/statcast_search?hfPT=&hfAB=&hfBBT=&hfPR=foul%7C&hfZ=&stadium=&hfBBL=&hfNewZones=&hfGT=R%7C&hfC=&hfSea=2019%7C&hfSit=&player_type=pitcher&hfOuts=&opponent=&pitcher_throws=&batter_stands=&hfSA=&game_date_gt=&game_date_lt=2019-06-05&hfInfield=&team=&position=&hfOutfield=&hfRO=&home_road=&hfFlag=&hfPull=&metric_1=&hfInn=&min_pitches=0&min_results=0&group_by=venue&sort_col=pitches&player_event_sort=h_launch_speed&sort_order=desc&min_pas=0#results) is an MLB affiliate that collects baseball statistics for major league games. The initial data set here `foul_balls` includes **906** observations with **7** variables. The dataset was used originally to observe where the most dangerous foul balls land, which is discussed in depth in [this story](https://fivethirtyeight.com/features/we-watched-906-foul-balls-to-find-out-where-the-most-dangerous-ones-land/). Just in my initial investigation, there appeared to be some `NA` data points, so we'll have to clean the data up.

**For this stage of analysis, I had to clean up the new dataset; although, I'll be keeping all of the variables for my analysis. These include:**

* __`matchup`__
    + The two teams that played. Formatted like `Atlanta Braves vs New York Mets`
* __`game_date`__
    + Date of the most foul-heavy day at each stadium. Formatted like `yyyymmdd`.
* __`type_of_hit`__
    + Fly, grounder, line drive, pop up or batter hit self
* __`exit_velocity`__
* __`predicted_zone`__
* __`camera_zone`__
* __`used_zone`__
