---
title: "Phase 2: Model Planning and Building"
author: "Brendan Allen"
output:
  html_document:
    df_print: paged
---

<style>
body {
text-align: justify}
</style>



#### [Home](https://ballen2024.github.io/DataScience/index.html)

#### [Phase 1: Discovery and Data Preparation](https://ballen2024.github.io/DataScience/deliverable1.html)

#### [Phase 2: Model Planning and Building](https://ballen2024.github.io/DataScience/deliverable2.html)

#### [Phase 3: Results and Operationalization](https://ballen2024.github.io/DataScience/deliverable3.html)

### Getting Started
**First we want to import any necessary libraries we will be using.**

**Note: Uncomment the install lines if you haven't yet installed these packages**
```{r warning=FALSE}
# install.packages("tidyr")
# install.packages("dplyr")
# install.packages("devtools")
# devtools::install_github("stefanedwards/lemon")
# install.packages("scales")
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("stringr")
# install.packages("caret")

suppressMessages(library(tidyr))
suppressMessages(library(dplyr))
suppressMessages(library(lemon))
suppressMessages(library(scales))
suppressMessages(library(tidyverse))
suppressMessages(library(knitr))
suppressMessages(library(stringr))
suppressMessages(library(caret))
knit_print.data.frame <- lemon_print
```

**We will want to import all of our work from [Phase 1](https://ballen2024.github.io/DataScience/deliverable1.html) as well.**
```{r  message=FALSE, error=FALSE, warning=FALSE, results='hide'}
purl("deliverable1.Rmd", output = "deliverable1.r")
source("deliverable1.r")
```
### Looking at the Data
The data source [**Baseball Savant**](https://baseballsavant.mlb.com/statcast_search?hfPT=&hfAB=&hfBBT=&hfPR=foul%7C&hfZ=&stadium=&hfBBL=&hfNewZones=&hfGT=R%7C&hfC=&hfSea=2019%7C&hfSit=&player_type=pitcher&hfOuts=&opponent=&pitcher_throws=&batter_stands=&hfSA=&game_date_gt=&game_date_lt=2019-06-05&hfInfield=&team=&position=&hfOutfield=&hfRO=&home_road=&hfFlag=&hfPull=&metric_1=&hfInn=&min_pitches=0&min_results=0&group_by=venue&sort_col=pitches&player_event_sort=h_launch_speed&sort_order=desc&min_pas=0#results) is an MLB affiliate that collects baseball statistics for major league games. The initial data set here `foul_balls` includes **906** observations with **7** variables. The dataset was used originally to observe where the most dangerous foul balls land, which is discussed in depth in [this story](https://fivethirtyeight.com/features/we-watched-906-foul-balls-to-find-out-where-the-most-dangerous-ones-land/). Our `launch_data` set was also collected from this source. I also used [**Baseball Reference**](https://www.baseball-reference.com/leagues/MLB/2016-standings.shtml) to gather data on standings for the 2016 MLB season. Just in my initial investigation, there appeared to be some `NA` data points, so we'll have to clean the data up. I also noticed that there is some inconsistent or inappropriate type formatting, so we'll have to clean that up as well.

#### Here is an overview of each dataset

##### `foul_balls`

* __`matchup`__
    + The two teams that played. Formatted like `Atlanta Braves vs New York Mets`. In order to have a merge point for this set and the previous set, I am going to make a lookup table with the team names and their corresponding three character codes.
* __`date`__
    + Date of the most foul-heavy day at each stadium. Formatted like `yyyy-mm-dd`.
* __`hit_type`__
    + Fly, grounder, line drive, pop up or batter hit self. Formated as a string.
* __`velocity`__
    + Recorded exit velocity of each hit -- blank if not provided. Formatted as a double
* __`predicted_zone`__
    + The zone we predicted the foul ball would land in by gauging angles. Formatted as a factor.
* __`camera_zone`__
    + The zone that the foul ball landed in, confirmed by footage. Formatted as a factor.
* __`used_zone`__
    + The zone used for analysis. Formatted as a factor.

##### `launch_data`

* __`date`__
    + Date of the game where this hit was recorded Formatted like `yyyy-mm-dd`.
* __`field`__
    + Field where the game was played. Formatted as string.
* __`home`__
    + Name of home team's 3 character abbreviation. Formatted as a string.
* __`away`__
    + Name of away team's 3 character abbreviation. Formatted as a string.
* __`velocity`__
    + Exit velocity of the ball hit. Formatted as a double.
* __`angle`__
    + Angle of the ball hit in degrees. Formatted as a double.

##### `standings`

* __`rank`__
    + Team rank overall. Formatted as a double
* __`team`__
    + Team name's 3 character abbreviation
* __`games`__
    + Total number of games played. Formatted as a double.
* __`wins`__
    + Total number of games won. Formatted as a double.
* __`losses`__
    + Total number of games lost. Formatted as a double.
* __`run_diff`__
    + Run differential (average runs scored - average runs allowed). Formatted as a double.


**Let's read in our new data on foul balls**
```{r, warning=FALSE}
foul_balls <- suppressMessages(read_csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/foul-balls/foul-balls.csv"))
launch_data <- suppressMessages(read_csv("./savant_data.csv"))
standings <- suppressMessages(read_csv("./standings.csv"))

```


**We'll want to clean up our data now**

In this block, I filtered our initial game data (`game_data`) to only include games from 2000-2016; omitted the missing data from our original foul balls data (`foul_balls`) from the article and our new foul ball data (`launch_data`) from the 2016 season; narrowed our data selection in all of our sets; and renamed the columns.
```{r, warning=FALSE}
game_data <- game_data %>%
  filter(date >= 19900101) %>%
  rename(away=v_name, home=h_name)

foul_balls <- foul_balls %>%
  na.omit() %>%
  rename(date = game_date, hit_type = type_of_hit, velocity = exit_velocity)

launch_data <- launch_data %>%
  select(c(game_date, player_name, home_team, away_team, launch_speed, launch_angle)) %>%
  filter(launch_speed != "null", launch_angle != "null") %>%
  rename(date=game_date, field=player_name, home=home_team, away=away_team, velocty=launch_speed, angle=launch_angle)

standings <- standings %>%
  select(c(Rk, Tm, G, W, L, W_L_per, Rdiff)) %>%
  rename(rank=Rk, team=Tm, games=G, wins=W, losses=L, win_loss_percentage= W_L_per, run_diff=Rdiff) %>%
  na.omit()
```

**Now we need to standardize the formatting of our variables**

Here we converted the dates to an actual date format, made our foul ball zones factors, and our velocity and angle to a numeric type instead of a character.
```{r warning=FALSE}
game_data$date <- as.character(game_data$date)
game_data$date <- as.Date(game_data$date, format="%Y%m%d")

foul_balls$predicted_zone <- as.factor(foul_balls$predicted_zone)
foul_balls$camera_zone <- as.factor(foul_balls$camera_zone)
foul_balls$used_zone <- as.factor(foul_balls$used_zone)

launch_data$velocty <- as.double(launch_data$velocty)
launch_data$angle <- as.double(launch_data$angle)

```

**To finalize cleaning our data, we need to do some work on the `matchup` column of our `foul_balls` data**

Specifically the matchup is formatted like `Atlanta Braves vs New York Mets`. We want to offer consistency with our other data sets, so we are going to split this `matchup` into `home` and `away`. We'll need to make a lookup table though to convert the full team name into it's 3 character abbreviation. I just created my own csv file for [this](https://github.com/ballen2024/ballen2024.github.io/blob/master/DataScience/teams.csv) in an editor, so we'll have to load that.
```{r warning=FALSE}
teams <- suppressMessages(read_csv("./teams.csv"))
```

Using regular expresions in `R`, we can split up the `matchup`'s into their `home` and `away` teams. We will want to grab everything in the string until we encounter "vs" or "VS" and store that as `away` team, since American convention holds that matchups are written as *away vs home*. We can throw away the "vs" or "VS" and then we can grab the second string and store it as the `home` team.

After a lot of trial an error to match the abbreviations to the full names in the `foul_balls` set, I finally realized the their was some unnoticed spaces at the front of the home teams and at the back of the away teams. This happened because of the way I used `str_split()`. I believe you could eat the whitespace, but I instead decided to add some spaces to the team names in the lookup table.

```{r warning=FALSE}
split <- str_split(foul_balls$matchup, "(vs|VS)")
mat <- matrix(unlist(split), ncol=2, byrow=TRUE)
matchup <- as.data.frame(mat)
foul_balls$home <- as.character(matchup$V2)
foul_balls$away <- as.character(matchup$V1)

teams$space_front <- paste("", teams$full)
teams$space_back <- paste(teams$full, "")
foul_balls$home <- teams[match(foul_balls$home, teams$space_front), "abb"]
foul_balls$away <- teams[match(foul_balls$away, teams$space_back), "abb"]
```

**Here is a quick look at our data after being cleaned up**

#### `game_data`
```{r, echo=FALSE, render=lemon_print}
head(game_data)
```

#### `foul_balls`
```{r, echo=FALSE, render=lemon_print}
head(foul_balls)
```

#### `launch_data`
```{r, echo=FALSE, render=lemon_print}
head(launch_data)
```

#### `standings`
```{r, echo=FALSE, render=lemon_print}
head(standings)
```

#### `teams`
```{r, echo=FALSE, render=lemon_print}
head(teams)
```



### Predictions to Model
Based on the initial phase and the newfound data, some predictions I would like to model are:

* Can you predict a team's record based on their run differential in the first month of the season?
    + If you could predict a team's record based on their run differential in the first month of the season, that would bring significant change to the climate of the first month of the season and I think there would be a difference in the trades occuring before the trade deadline in July. I will be exploring a win-loss percentage rather than an exact record at this time.
* Do teams that hit more homeruns win more games?
    + If it was found that teams that hit more homeruns win more games, teams would probably want to draft more heavy hitters. Perhaps a look at which factors are significant in a teams record would be a good model.


### Model Planning

**First I'll work on a model for predicting a team's win-loss percentage based on their run differential in the first month of the season**

To do this, I created a tibble `run_differntial` with the data I wanted.  I filtered my data to only include games from the first month of the 2016 season. I then created a new column for the home team's run differential for that game as well as for the away team.

I then created another tibble named `model`. This is what we will be training and testing our model with.


```{r}
run_differential <- tibble(date=game_data$date, 
                           home=game_data$home, 
                           away=game_data$away, 
                           h_score=game_data$h_score, 
                           v_score=game_data$v_score)
run_differential <- run_differential %>%
  filter(date >= "2016-04-03") %>%
  filter(date <= "2016-05-03")

run_differential <- run_differential %>%
  mutate(h_run_diff=(h_score-v_score), v_run_diff=(v_score-h_score))

# I had to use the following 2 lines to get the win-loss percentage as a vector rather than a dataframe.
# It seems when I extracted the match it stored it as a datafrane even though it was a list, which made
# accessing it a pain, so I decided to strip it down to a vector first
win_loss <- standings[match(teams$abb, standings$team), "win_loss_percentage"]
win_loss <- as.vector(win_loss$win_loss_percentage)

model <- tibble(team=teams$abb,
                run_diff=0,
                win_loss_percentage=win_loss
                )

# I realize that this is a very inefficient algorithm, but for my purpose of only analyzing less 
# than 400 x 30 observations, I deemed it okay. I'm sure there is a way to leverage the vectorization
# of R to do this much more efficiently.
for (x in 1:length(run_differential$date)) {
  for (t in 1: length(model$team)) {
    if (model$team[t] == run_differential$away[x]) {
      model$run_diff[t] <- model$run_diff[t] + run_differential$v_run_diff[x]
    }
    if (model$team[t] == run_differential$home[x]) {
      model$run_diff[t] <- model$run_diff[t] + run_differential$h_run_diff[x]
    }
  }
}
```

**Now we are ready to create our model**

```{r}
set.seed(1) # set the seed to yield consistent values in the data partition
selection <- createDataPartition(model$win_loss_percentage, p=0.75, list=FALSE)

train <- model[selection, ]
test <- model[-selection, ]


train_model <- lm(data = train, formula=win_loss_percentage~run_diff)

predictions <- train_model %>% 
  predict(test)
```

In summarizing our model, we can see that the run differential in the first month of the season is a good predictor of a team's win-loss percentage. We can see this because our p-value is very small. Additionally we can predict that for every 1 run increase in the run differneital, we will have a 0.0015 point increase in the win-loss percentage of that team.
```{r}
summary(train_model)
summary(predictions)
```

We can see that our root-mean squared error method of measuring our model performance is best and shows that we have a `r round(RMSE(predictions, test$win_loss_percentage)/mean(test$win_loss_percentage), 4)` error rate.
```{r, render=lemon_print}
data.frame(R2 = R2(predictions, test$win_loss_percentage),
            RMSE = RMSE(predictions, test$win_loss_percentage),
            MAE = MAE(predictions, test$win_loss_percentage))

```

The following plot shows that we have a pretty linear relationship between our predicted win-loss percentage values and the actual values, which means our model works well, especially considering that we only pulled from the first month of the season.
```{r}
ggplot(test, aes(x=predictions, y=win_loss_percentage)) + geom_point() + labs(x="Predicted Win-Loss Percentage",y="Actual Win-Loss Percentage",title="Model Predictions vs. Actual Values") + geom_abline()
```

**For this next model, we'll explore if teams that hit more homeruns win more games.**
The motivation for modeling this comes from the discussion that "small ball" is better for winning games. However, if we can see a significant connection between the number of homeruns a teams hits and the games they win, perhaps this discussion can be put to rest.

I'll again be looking at the 2016 season, using `game_data` to collect the homeruns hit by a team and `standings` to collect the wins for each team.

```{r}
homeruns <- tibble(date=game_data$date,
                   home=game_data$home,
                   away=game_data$away,
                   h_homeruns=game_data$h_homeruns,
                   v_homeruns=game_data$v_homeruns)


wins <- standings[match(teams$abb, standings$team), "wins"]
wins <- as.vector(wins$wins)

homeruns_model <- tibble(team=teams$abb,
                homeruns=0,
                wins=wins
                )

for (x in 1:length(homeruns$date)) {
  for (t in 1: length(homeruns_model$team)) {
    if (homeruns_model$team[t] == homeruns$away[x]) {
      homeruns_model$homeruns[t] <- homeruns_model$homeruns[t] + homeruns$v_homeruns[x]
    }
    if (homeruns_model$team[t] == homeruns$home[x]) {
      homeruns_model$homeruns[t] <- homeruns_model$homeruns[t] + homeruns$v_homeruns[x]
    }
  }
}
```

